\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata} \textbackslash{} Appendix

### 2023-07-28 manuscript_v7/01-background.qmd
updated conclusion and discussion

Reasons given for this inequality in vaccination uptake have included medical mistrust due to systemic medical racism [@stoler2021;@mosby2021], mistrust in vaccines [@willis2021], and the existence of health misinformation and disinformation [@bogart2021;@mosby2021;@freeman2020]. However, it is important to consider that vaccination uptake can also vary geographically across the administrative or political subdivisions within a country. For example, differences in COVID-19 vaccination rates have been associated with differences in attitudes towards vaccination in administrative areas in New York and Chicago[@malik2020], scarcity of vaccination facilities in areas where underrepresented groups reside in Toronto[@bogoch2022], as well as an absence of prioritization of areas inhabited by vulnerable groups in the southeastern region of the US[@nguyen2021]. Other studies have also shown heterogeneity in vaccine uptake within counties in the US [@mollalo2021; @yang2022; @tiu2022; @bhuiyan2022], indicating that accounting for geographical differences in vaccination can help predict patterns of booster uptake in Scottish communities [@wood2022]. Therefore, the study of vaccination uptake across administrative of political areas can be useful to highlight differences that governments or public health agencies might need to address. However, in the case of Canada, there is a limited amount of studies that have analyzed geographical differences in vaccination. Existing studies in this area have focused on differences within certain cities, such as Toronto [@choi2021], or Montréal [@mckinnon2021]), or have explored differences between provinces-at-large [@guay2022], but to our knowledge, there are no studies that analyze differences in vaccination uptake at the intra-provincial level. Such need is specially important in the context of Canada's pandemic response goals, which have been to minimize serious illness and deaths while minimizing societal disruption [@naci2023]. Analyzing differences in vaccination uptake within the provinces can aid to identify inequalities that might exist and that need to be addressed before the advent of another pandemic.

>>>>

Reasons given for this inequality in vaccination uptake have included medical mistrust due to systemic medical racism [@stoler2021;@mosby2021], mistrust in vaccines [@willis2021], and the existence of health misinformation and disinformation [@bogart2021;@mosby2021;@freeman2020]. However, it is important to consider that vaccination uptake can also vary geographically across the administrative or political subdivisions within a country. For example, differences in COVID-19 vaccination rates have been associated with differences in attitudes towards vaccination in administrative areas in New York and Chicago[@malik2020], scarcity of vaccination facilities in areas where underrepresented groups reside in Toronto when the provincial vaccination program started[@bogoch2022], as well as an absence of prioritization of areas inhabited by vulnerable groups in the southeastern region of the US[@nguyen2021]. Other studies have also shown heterogeneity in vaccine uptake within counties in the US [@mollalo2021; @yang2022; @tiu2022; @bhuiyan2022], indicating that accounting for geographical differences in vaccination can help predict patterns of booster uptake in Scottish communities [@wood2022]. Therefore, the study of vaccination uptake across administrative of political areas can be useful to highlight differences that governments or public health agencies might need to address. However, in the case of Canada, there is a limited amount of studies that have analyzed geographical differences in vaccination. Existing studies in this area have focused on differences within certain cities, such as Toronto [@choi2021], or Montréal [@mckinnon2021]), or have explored differences between provinces-at-large [@guay2022], but to our knowledge, there are no studies that analyze differences in vaccination uptake at the intra-provincial level. Such need is specially important in the context of Canada's pandemic response goals, which have been to minimize serious illness and deaths while minimizing societal disruption [@naci2023]. Analyzing differences in vaccination uptake within the provinces can aid to identify inequalities that might exist and that need to be addressed before the advent of another pandemic.

### 2023-07-28 manuscript_v7/04-discussion.qmd
updated conclusion and discussion

Furthermore, there is an ongoing challenge for the health system of the province with regard to personalized healthcare for marginalized individuals. For example, the West Health Region has only two Aboriginal Health Access Centres (community-led primary healthcare organizations focused on First Nations, Métis, and Inuit communities) to provide care to an estimated population of 100,000 Indigenous individuals living in the area [@ontariohealth]. Lack of access to personalized healthcare affects individuals that may mistrust the traditional healthcare system due to systemic racism or oppression, which is known to be the case for Indigenous and Black individuals in Canada, as these rationales have been associated with observed lower vaccination rates among these groups [@smylie2022;@eissa2021]. Taken together, this suggests that healthcare disparities specific to these underrepresented groups in certain parts of the province impacted vaccination uptake, and highlights the need of investments in the Health Regions focused on resources, infrastructure, and specially personnel that can deliver personalized care to marginalized communities, as it has been shown that such efforts have improved trust in vaccination in underrepresented groups elsewhere [@schafferderoo2020].

>>>>

Furthermore, there is an ongoing challenge for the health system of the province with regard to personalized healthcare for marginalized individuals. Indeed, previous studies have pointed out mistrust in the traditional healthcare system (due to systemic racism or oppression) as a rationale for lower vaccination rates in the case of Indigenous and Black individuals in Canada [@smylie2022;@eissa2021]. Other studies have also shown a need for increased culturally responsive care for other underrepresented groups such as Black women and and other immigrant groups in Canada and Ontario [@etowa2021;@patel2019]. Considering that for example, the West Health Region has only two Aboriginal Health Access Centres (community-led primary healthcare organizations focused on First Nations, Métis, and Inuit communities) to provide care to an estimated population of 100,000 Indigenous individuals living in the area [@ontariohealth], it is possible that the limited personalized healthcare for underrepresented groups in certain parts of the province impacted vaccination uptake as well, and highlights the need of investments in the Health Regions focused on resources, infrastructure, and specially personnel that can deliver personalized care to marginalized communities, as it has been shown that such efforts have improved trust in vaccination in underrepresented groups elsewhere [@schafferderoo2020].

### 2023-07-28 manuscript_v7/05-conclusion.qmd
updated conclusion and discussion

The evidence collected in this study shows that differences in the uptake of the first dose of the COVID-19 vaccine by underrepresented individuals in the province of Ontario were the result of complex relationships between infrastructure, resources, occupations, geography, and socio-economic factors. Specifically, we identified that between the last trimester of 2021 and January of 2022 individuals that identified as Arab or Middle Eastern, Black, Latin American, of a mixed background, or that belonged to other racial groups had significantly lower odds of vaccination when compared to White/Caucasian individuals in Ontario.

However, although overall vaccination uptake was lower in these underrepresented individuals in the province as a whole, the odds of vaccination differed within certain income levels and Health Regions in these groups, specially for those that reported a low household income and that had higher vaccination odds that individuals in the same racial/ethnic group with a higher income. These results highlight the complex landscape of the province, where varying degrees of rurality exist in conjunction with a socio-demographic makeup that is unique to each of the Health Regions.

Such complexity presents an ongoing challenge for the healthcare system of Ontario, and in particular, to the Health Region model which will need to adapt and build on the LHIN model with the aim of providing an integrative approach that can serve those that are most vulnerable across all the Health Regions of the province. In particular, personalized care is an area that could be further developed to improve vaccination uptake in underrepresented individuals in the future. As we have shown above, there are currently a limited number of centers focused on community-led healthcare in the province, and our results showed the existence of differences in vaccination uptake between certain Health Regions. Improving personalized care for underrepresented individuals then, can serve as a point to improve trust and facilitate vaccine access in these marginalized communities.

<!-- The implementation of the Health Regions in Ontario aimed at reducing the bureaucratic complexity and health disparities identified under the LHIN model. However, there are certain challenges that this new model will need to overcome in order to work syngergistically with the rest of the healthcare system of the province. First, each Health Region now covers a large geographical area that was served in the past by multiple LHINs. This in turn creates a complex socio-demographic landscape within each Health Region that is likely to be different in each case due the different levels of rurality and the proportion of underrepresented groups in each Health Region.  -->

Furthermore, in the event of a future pandemic or public health emergency in the province in the years ahead, PHO and OH will need to collaborate on the implementation of public health measures (such as vaccination campaigns) building on the experience gained during the COVID-19 vaccination rollout. So far, the COVID-19 pandemic served as a stress test for both agencies, showing the necessity from decision makers to clearly define how PHO and OH will interact in the future in order to ensure that Ontarians can benefit by virtue of a robust collaboration between both agencies.

<!-- The evidence collected in this study shows that healthcare disparities experienced by underrepresented individuals in the province are the result of complex relationships between infrastructure, resources, occupations, geography, and socio-economic factors, among others. Therefore, this complexity presents an ongoing challenge for the healthcare system of Ontario, which will need to adapt and continue to aim for an integrative approach that can serve those that are most vulnerable. -->

<!-- So far, the evidence collected during the COVID-19 pandemic indicates that differences in vaccination uptake are associated to a lack of infrastructure and resources that can adequately support personalized care to marginalized individuals. In the near future, health decision-makers will need to consider the implementation of policies that are focused on addressing this problematic. -->

Finally, the recent nature in the adaption of the Health Region poses a challenge for researchers in the acquisition of data and information that can be used to analyze the performance of the new model. From one side, the Health Regions have not been incorporated as part of Census data (LHINs were considered before in the Census), and this impacts the amount of data that can be obtained. On the other hand, evaluations of the Health Region model are at the moment not considered in the Annual Reports of the Auditor General of Ontario, which was critical in the past to identify limitations in the LHIN model. Currently, the only demographic information available for each Health Region is provided by Ontario Health (the agency that administers the Health Regions) but this information only provides general estimates that do not allow for detailed analyses of performance indicators (such as hospitalizations, readmissions, and trends in chronic disease incidence) between the Regions. Therefore, there is a pressing need for open data that can be used by researchers and decision-makers to assess the performance of the Health Region model.

<!-- The Health Region model will only by successful if it ensures that healthcare improves across all segments of the population of Ontario, particularly in the event of a future public health emergency or pandemic where so far, based on the experience of the COVID-19 pandemic, underrepresented individuals have been disproportionately affected. -->

>>>>

The evidence collected in this study shows differences in the uptake of the first dose of the COVID-19 vaccine in Ontario between October 2021 and January 2022 in underrepresented groups, which had significantly lower odds of vaccination when compared to White/Caucasian individuals. However, although overall vaccination uptake was lower in underrepresented individuals in the province as a whole, the odds of vaccination differed within certain income levels and Health Regions in these groups. Specifically, those that reported a low household income had higher vaccination odds when compared to individuals in the same racial/ethnic group in a higher income bracket. These results highlight the complex landscape of the province, where varying degrees of rurality exist in conjunction with a socio-demographic makeup that is unique to each of the Health Regions.

Personalized care is an area that could be further developed to improve vaccination uptake in underrepresented individuals in the future. As we have shown in the discussion, there are currently a limited number of centers focused on community-led healthcare in the province. Our results showed the existence of differences in vaccination uptake between certain Health Regions and therefore, improving personalized care for underrepresented individuals can serve as a point to improve trust and facilitate vaccine access in these marginalized communities.

Overall, there are ongoing challenges for healthcare access across different segments of the population of Ontario. Future studies can be focused at more in-depth analyses of vaccination uptake within the different socio-demographic groups in each Health Region, in order to provide decision-makers with information that can serve to carefully consider and address how to improve vaccine access to marginalized communities in Ontario in the event of a future public health emergency.

### 2023-07-28 manuscript_v7/appendix.qmd
updated conclusion and discussion

---
title: '**Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata** \\ Appendix'
author:
  - name: Ariel Mundo Ortiz
    affil-id: 1,2,3
  - name: Bouchra Nasri
    affil-id: 1,2,3,\*
affiliations: 
  - id: 1
    name: Centre de Recherches Mathématiques, University of Montreal, Montréal, Canada
  - id: 2
    name: Department of Social and Preventive Medicine, École de Santé Publique, University of Montreal, Montréal, Canada
  - id: 3
    name: Centre de recherche en santé publique, University of Montreal, Montréal, Canada
  - id: \*
    name: Corresponding author, \href{mailto:bouchra.nasri@umontreal.ca}{bouchra.nasri@umontreal.ca}
format: 
  pdf:
    keep-tex: true
    template-partials: 
      - latex/title.tex
    include-in-header:
      - text: |
         \usepackage[noblocks]{authblk}
         \renewcommand*{\Authsep}{, }
         \renewcommand*{\Authand}{, }
         \renewcommand*{\Authands}{, }
         \renewcommand\Affilfont{\small}
      - file: latex/appendix_preamble.sty
linestretch: 2
---

>>>>

---
title: '**Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata** \\ Appendix'
author:
  - name: Ariel Mundo Ortiz
    affil-id: 1,2,3
  - name: Bouchra Nasri
    affil-id: 1,2,3,\*
affiliations: 
  - id: 1
    name: Centre de Recherches Mathématiques, University of Montreal, Montréal, Canada
  - id: 2
    name: Department of Social and Preventive Medicine, École de Santé Publique, University of Montreal, Montréal, Canada
  - id: 3
    name: Centre de recherche en santé publique, University of Montreal, Montréal, Canada
  - id: \*
    name: Corresponding author, \href{mailto:bouchra.nasri@umontreal.ca}{bouchra.nasri@umontreal.ca}
format: 
  pdf:
    geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm
    keep-tex: true
    template-partials: 
      - latex/title.tex
    include-in-header:
      - text: |
         \usepackage[noblocks]{authblk}
         \renewcommand*{\Authsep}{, }
         \renewcommand*{\Authand}{, }
         \renewcommand*{\Authands}{, }
         \renewcommand\Affilfont{\small}
      - file: latex/appendix_preamble.sty
#linestretch: 2
---

### 2023-07-28 manuscript_v7/appendix.tex
updated conclusion and discussion

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi

\setstretch{2}
This document provides details on assessing the socio-economic
information from the clean survey dataset, and how it compared to the
Census Data for ontario. At the end, it also provides information about
the corrections implemented into the data. The variables examined were:
race/ethnicity, age groups, income, and population by Health Region. The
computations are performed in the \texttt{clean\_dataset.csv} file that
was created at the end of the data cleaning process. Population-wide
totals were obtained from the 2016, available in the
\href{https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/prof/details/Page.cfm?Lang=E\&Geo1=PR\&Code1=35\&Geo2=\&Code2=\&Data=Count\&SearchText=Ontario\&Sear}{Statistics
Canada Website}.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=20mm,left=20mm,right=20mm,bottom=20mm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

This document provides details on assessing the socio-economic
information from the clean survey dataset, and how it compared to the
Census Data for ontario. At the end, it also provides information about
the corrections implemented into the data. The variables examined were:
race/ethnicity, age groups, income, and population by Health Region. The
computations are performed in the \texttt{clean\_dataset.csv} file that
was created at the end of the data cleaning process. Population-wide
totals were obtained from the 2016, available in the
\href{https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/prof/details/Page.cfm?Lang=E\&Geo1=PR\&Code1=35\&Geo2=\&Code2=\&Data=Count\&SearchText=Ontario\&Sear}{Statistics
Canada Website}.
